---
title: "Bayesian Statistics in Social Sciences"
subtitle: "Class 4 - analysis and Plotting of MCMC Draws"
author: "Wiktor Soral"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo=FALSE, 
  warning=FALSE, 
  message = FALSE,
  fig.align='center', 
  fig.dpi=300, 
  out.width='65%'
)
```

```{r}
library(tidyverse)
library(bayesplot)
theme_set(theme_classic())
```

## Finding denominator of Bayes theorem

- Obtaining the posterior distribution requires computing the denominator.
- The result of the integration is a constant used to normalize the posterior distribution.

$$
P(\theta | \mathcal{D}) = \frac{P(\mathcal{D} | \theta)\times P(\mathcal{D} | \theta)}{\int{P(\mathcal{D} | \theta)\times P(\mathcal{D} | \theta)}d\theta}
$$

---

## Function

```{r}
tibble(x = seq(-4, 4, length.out = 100)) %>% 
  ggplot(aes(x = x))+
  stat_function(fun=dnorm, size=1.1)+
  labs(x=expression(theta), y=expression("P("~theta~")"))
```

---

## Finding area under function

```{r}
tibble(x = seq(-4, 4, length.out = 100)) %>% 
  ggplot(aes(x = x))+
  stat_function(fun=dnorm, geom="polygon", colour="black", fill="lightblue", size=1.1)+
  labs(x=expression(theta), y=expression("P("~theta~")"))
```

---

## Approximating area under function

```{r}
rectangle <- 
  tibble(x = c(-4, 4, 4, -4),
         y = c(0, 0, 0.4, 0.4))

tibble(x = seq(-4, 4, length.out = 100)) %>% 
  ggplot(aes(x = x))+
  stat_function(fun=dnorm, size=1.1)+
  labs(x=expression(theta), y=expression("P("~theta~")"))+
  geom_polygon(mapping = aes(x=x, y=y), data=rectangle, colour="red", fill=NA, linetype =2)
```

---

## Approximating area under function

```{r}
sims <- tibble(
  x_sim = runif(5000, -4, 4),
  y_sim = runif(5000, 0, 0.4)
  ) %>% 
  mutate(accept = ifelse(y_sim <= dnorm(x_sim), "yes", "no"))

tibble(x = seq(-4, 4, length.out = 100)) %>% 
  ggplot(aes(x = x))+
  stat_function(fun=dnorm, size=1.1)+
  labs(x=expression(theta), y=expression("P("~theta~")"))+
  geom_polygon(mapping = aes(x=x, y=y), data=rectangle, colour="red", fill=NA, linetype =2)+
  geom_point(aes(x=x_sim, y=y_sim), data=sims, alpha=.7)
```

---

## Approximating area under function

```{r}
tibble(x = seq(-4, 4, length.out = 100)) %>% 
  ggplot(aes(x = x))+
  stat_function(fun=dnorm, size=1.1)+
  labs(x=expression(theta), y=expression("P("~theta~")"))+
  geom_polygon(mapping = aes(x=x, y=y), data=rectangle, colour="red", fill=NA, linetype =2)+
  geom_point(aes(x=x_sim, y=y_sim, colour=accept), data=sims, alpha=0.7)+
  guides(colour="none")
```

---

## Approximating area under function

Accepted values of x follow a Normal distribution.

```{r}
sims %>% 
  filter(accept=="yes") %>% 
  ggplot(aes(sample=x_sim))+
  stat_qq(alpha = 0.4,)+
  stat_qq_line()+
  labs(x="Normal theoretical quantiles", y= "Sample quantiles")
```

---

## Markov chain Monte Carlo

- MCMC refers to a class of algorithms based on a shared idea:
  - Metropolis, Metropolis–Hastings, Gibbs
  - Hamiltonian Monte Carlo, NUTS
- The goal is to sample from a complex, high-dimensional probability distribution.
- This is done by repeatedly generating candidate samples from a proposal distribution and accepting or rejecting them according to a specified acceptance probability.

---

## How MCMC explores the posterior

- Imagine a random walk across possible parameter values
- We start at some initial point
- At each step, we propose a small move to a nearby value
- If the new value is more plausible, we are likely to move there
- If it is less plausible, we might still move there—but less often
- Over time, we spend more time in regions with higher probability
- Early steps are unstable, so we discard them (warm-up / burn-in)
- The remaining steps trace out the shape of the posterior distribution

---

## Markov chain Monte Carlo

```{r}
library(bayesplot)
x <- example_mcmc_draws(1, 1)
mcmc_trace(x)+
  ggtitle("A traceplot of 250 draws from a posterior distribution")
```

---

## Markov chain Monte Carlo

```{r}
mcmc_hist(x)+
  ggtitle("A histogram of 250 draws from a posterior distribution")
```

---

## Markov chain Monte Carlo

```{r}
mcmc_dens(x)+
  ggtitle("A density plot of 250 draws from a posterior distribution")
```

---
class: inverse, center, middle

# Describing posterior distribution

---
## Centrality - mean

```{r}
mcmc_dens(x)+
  geom_vline(xintercept = -21.06, colour="red")+
  ggtitle("A density plot of 250 draws from a posterior distribution",
          subtitle = "with the red line denoting posterior mean")
```

---

## Centrality - MAP estimate

```{r}
mcmc_dens(x)+
  geom_vline(xintercept = -24.17, colour="green")+
  ggtitle("A density plot of 250 draws from a posterior distribution",
          subtitle = "with the green line denoting the maximum a posteriori (MAP) estimate")
```

---

## Centrality - median

```{r}
mcmc_dens(x)+
  geom_vline(xintercept = -22.14, colour="blue")+
  ggtitle("A density plot of 250 draws from a posterior distribution",
          subtitle = "with the blue line denoting the median")
```

---

## Uncertainty - credible intervals

```{r}
mcmc_dens(x)+
  geom_polygon(data=tibble(Value=c(-46.58, 10.13, 10.13, -46.58), y=c(0,0,0.029,0.029)), aes(y=y), fill="yellow", alpha=1/3)+
  ggtitle("A density plot of 250 draws from a posterior distribution",
          subtitle = "with the yellow area denoting the equal-tailed 95% credible interval [-46.58, 10.13]")
```

---

## Uncertainty - highest density intervals

```{r}
mcmc_dens(x)+
  geom_polygon(data=tibble(Value=c(-46.70, 10.31, 10.31, -46.70), y=c(0,0,0.029,0.029)), aes(y=y), fill="cyan", alpha=1/3)+
  ggtitle("A density plot of 250 draws from a posterior distribution",
          subtitle = "with the cyan area denoting the 95% highest density interval [-46.70, 10.31]")
```

---

## Uncertainty - highest density intervals (multimodal)

```{r}
mcmc_dens(rbind(x, x+100))+
  geom_polygon(data=tibble(Value=c(-56.89950 , 15.1946, 15.1946, -56.89950), y=c(0,0,0.0104,0.0104)), aes(y=y), fill="cyan", alpha=1/3)+
  geom_polygon(data=tibble(Value=c(42.59036 , 115.0449, 115.0449, 42.59036), y=c(0,0,0.0104,0.0104)), aes(y=y), fill="cyan", alpha=1/3)+
  ggtitle("A density plot of 500 draws from a posterior distribution",
          subtitle = "with the cyan areas denoting the 95% highest density intervals [-56.90, 15.19] and [42.59, 115.05]")
```

---

## Existence - $P_{D}$

```{r}
library(bayestestR)
fit <- readRDS("../notebooks/data_files/c2_fit.rds")

plot(p_direction(fit))
```

- Probability of Direction: 0.93

---

## Existence - $P_{D}$


- pd \<= 95%: uncertain
- pd \> 95%: possibly existing
- pd \> 97%: likely existing
- pd \> 99%: probably existing
- pd \> 99.9%: certainly existing


---

## Significance - $ROPE$

```{r}
rope(x, range=c(-10,10)) %>% 
  plot()+
  theme_classic()
```

-   Proportion of samples inside ROPE: 22.88%

---

## Significance - $ROPE$


- more than 99% in ROPE: negligible (we can accept the null hypothesis)
- less than 97.5% in ROPE: probably negligible
- \<=97.5% & \>= 2.5% in ROPE: undecided significance
- \<2.5% in ROPE: probably significant
- \<1% in ROPE: significant (we can reject the null hypothesis)
