---
title: "Bayesian Statistics in Social Sciences"
subtitle: "Class 6 - Model Comparison and Averaging"
author: "Wiktor Soral"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo=FALSE, 
  warning=FALSE, 
  message = FALSE,
  fig.align='center', 
  fig.dpi=300, 
  out.width='65%'
)

```

## Model comparison - Bayes factor 

-   Bayes Factor (BF) indicates the relative strength of evidence in the data for two competing theories/models/hypotheses
    -   Example: Alternative hypothesis ($H_1$) vs. Null hypothesis ($H_0$)
-   $BF$ ranges from 0 to $+\infty$
-   $BF_{10}$ to indicate how much more likely the alternative hypothesis compared to the null hypothesis
-   Values $> 1$ support the alternative hypothesis, whereas values $< 1$ support the null hypothesis
-   $BF_{01}$ to indicate how much more likely the null hypothesis compared to the alternative hypothesis
-   $BF_{01} = 1/BF_{10}$

---

## Model comparison - Bayes factor

|       | BF  |       | Interpretation                 |
|-------|-----|-------|--------------------------------|
| 100   | \<  |       | Extreme evidence for $H_1$     |
| 30    | \-  | 100   | Very strong evidence for $H_1$ |
| 10    | \-  | 30    | Strong evidence for $H_1$      |
| 3     | \-  | 10    | Substantial evidence for $H_1$ |
| 1     | \-  | 3     | Anecdotal evidence for $H_1$   |
|       | 1   |       | No evidence                    |
| 1/3   | \-  | 1     | Anecdotal evidence for $H_0$   |
| 1/10  | \-  | 1/3   | Substantial evidence for $H_0$ |
| 1/30  | \-  | 1/10  | Strong evidence for $H_0$      |
| 1/100 | \-  | 1/30  | Very strong evidence for $H_0$ |
|       | \>  | 1/100 | Extreme evidence for $H_0$     |

---

## Overfitting

```{r}
library(rcartocolor)
library(tidyverse)
library(patchwork)
d <- 
  tibble(species = c("afarensis", "africanus", "habilis", "boisei", "rudolfensis", "ergaster", "sapiens"), 
         brain   = c(438, 452, 612, 521, 752, 871, 1350), 
         mass    = c(37.0, 35.5, 34.5, 41.5, 55.5, 61.0, 53.5))
p <-
  d %>% 
  ggplot(aes(x = mass, y = brain)) +
  geom_point(color = carto_pal(7, "BurgYl")[7]) +
  scale_x_continuous("body mass (kg)", limits = c(33, 62), expand = c(0, 0)) +
  ylab("brain volume (cc)") +
  coord_cartesian(ylim = c(300, 1500)) +
  theme_classic() +
  theme(text = element_text(family = "Courier"),
        panel.background = element_rect(fill = alpha(carto_pal(7, "BurgYl")[3], 1/4)))

# linear
p1 <- 
  p +
  stat_smooth(method = "lm", fullrange = TRUE, level = .89,  # note our rare use of 89% intervals
              color = carto_pal(7, "BurgYl")[6], fill = carto_pal(7, "BurgYl")[6], 
              linewidth = 1/2, alpha = 1/3,
              formula = y ~ x) +
  ggtitle(NULL, subtitle = expression(italic(R)^2==".49"))
  
# quadratic
p2 <-
  p + 
  stat_smooth(method = "lm", fullrange = TRUE, level = .89,
              color = carto_pal(7, "BurgYl")[6], fill = carto_pal(7, "BurgYl")[6], 
              linewidth = 1/2, alpha = 1/3,              
              formula = y ~ poly(x, 2)) +
  ggtitle(NULL, subtitle = expression(italic(R)^2==".54"))

# cubic
p3 <-
  p + 
  stat_smooth(method = "lm", fullrange = TRUE, level = .89,
              color = carto_pal(7, "BurgYl")[6], fill = carto_pal(7, "BurgYl")[6], 
              linewidth = 1/2, alpha = 1/3,              
              formula = y ~ poly(x, 3)) +
  ggtitle(NULL, subtitle = expression(italic(R)^2==".68"))

# fourth-order polynomial
p4 <-
  p + 
  stat_smooth(method = "lm", fullrange = TRUE, level = .89,
              color = carto_pal(7, "BurgYl")[6], fill = carto_pal(7, "BurgYl")[6], 
              linewidth = 1/2, alpha = 1/3,              
              formula = y ~ poly(x, 4)) +
  ggtitle(NULL, subtitle = expression(italic(R)^2==".81"))

# fifth-order polynomial
p5 <-
  p + 
  stat_smooth(method = "lm", fullrange = TRUE, level = .89,
              color = carto_pal(7, "BurgYl")[6], fill = carto_pal(7, "BurgYl")[6], 
              linewidth = 1/2, alpha = 1/3,              
              formula = y ~ poly(x, 5)) +
  # we're adjusting the y-axis range for this plot (and the next)
  coord_cartesian(ylim = c(150, 1900)) +
  ggtitle(NULL, subtitle = expression(italic(R)^2==".99"))
  
# sixth-order polynomial
p6 <-
  p + 
  # mark off 0 on the y-axis
  geom_hline(yintercept = 0, color = carto_pal(7, "BurgYl")[2], linetype = 2) + 
  stat_smooth(method = "lm", fullrange = TRUE, level = .89,
              color = carto_pal(7, "BurgYl")[6], fill = carto_pal(7, "BurgYl")[6], 
              linewidth = 1/2, alpha = 1/3,
              formula = y ~ poly(x, 6)) +
  coord_cartesian(ylim = c(-300, 1500)) +
  ggtitle(NULL, subtitle = expression(italic(R)^2==1))

(p1 + p2) / (p3 + p4) / (p5 + p6)
```


---

## Overfitting

```{r}
dev_tibble <- read_csv2("figures/dev_tibble.csv")
# this intermediary tibble will make `geom_text()` easier
dev_text <-
  dev_tibble %>% 
  filter(n_par > 1.5, 
         n_par < 2.5) %>% 
  mutate(n_par = ifelse(sample == "in", n_par - .25, n_par + .33))
  
# the plot
dev_tibble %>% 
  ggplot(aes(x = n_par, y = mean,
             ymin = mean - sd, ymax = mean + sd,
             group = sample, color = sample, fill  = sample)) +
  geom_pointrange(shape = 21) +
  geom_text(data = dev_text,
            aes(label = sample),
            family = "Courier") +
  scale_color_manual(values = c(carto_pal(7, "BurgYl")[7], carto_pal(7, "BurgYl")[5])) +
  scale_fill_manual(values  = c(carto_pal(7, "BurgYl")[5], carto_pal(7, "BurgYl")[7])) +
  labs(x = "number of parameters",
       y = "deviance") +
  theme_classic() +
  theme(text = element_text(family = "Courier"),
        legend.position = "none",
        panel.background = element_rect(fill = alpha(carto_pal(7, "BurgYl")[3], 1/4)),
        strip.background = element_rect(fill = alpha(carto_pal(7, "BurgYl")[1], 1/4), color = "white"))
```




---

## Leave-one-out cross validation

```{r}
set.seed(1234)
y = rnorm(6)
tibble(x = rep(1:6, 6),
       y = rep(y, 6),
       step = rep(1:6, each = 6)) %>% 
  mutate(leave_one = ifelse(x == step, "test", "train")) %>% 
  ggplot(aes(x, y))+
  facet_wrap(~step)+
  geom_point(aes(colour=leave_one))+
  scale_colour_manual(values = c("gray","black"))+
  labs(colour="")+
  theme_classic()
  
```

---

## ELPD and LOOIC

$$p(y_i | y_{-i})$$
--

$$elpd_i = log\ p(y_i | y_{-i})$$
--

$$elpd_{loo} = \sum_{i=1}^n  log\ p(y_i | y_{-i})$$

--

$$LOOIC = -2 * elpd_{loo}$$

---

## PSIS-LOOIC

$$log\ p(y_i | y_{-i}) \approx log \left( \frac{\sum_s w_i^{(s)} p(y_i | \theta^{(s)})}{\sum_s w_i^{(s)}} \right)$$
where:

$$w^{(s)} = \frac{1}{p(y_i | \theta^{(s)})}$$

- Some weigths $w^{(s)}$ may be very large and lead to noisy estimator.
- PSIS-LOOIC uses Pareto distribution fitted to the tail of weights to smooth very large values.
- It may not always work properly. See Pareto $\hat{k}$ parameter
  - $\hat{k}$ < 0.5 - great
  - $\hat{k}$ < 0.7 - OK
  - $\hat{k}$ > 0.7 - approximate LOO may not work properly
  
---

## Effective Number of Parameters in PSIS-LOO ($p_{\text{loo}}$)

**Definition**  
- $p_{\text{loo}}$ = effective number of parameters  
- Measures how much the model adapts to the observed data  

**Intuition**  
- Quantifies **model flexibility / complexity in practice**  
- Not equal to nominal parameter count  
- Reflects **overfitting potential**  


**Computation (conceptually)**  
- Difference between:
  - fit to observed data  
  - fit under leave-one-out prediction  
- $p_{\text{loo}} = \sum_i \big(\log p(y_i \mid y) - \log p(y_i \mid y_{-i})\big)$ 

---

## Effective Number of Parameters in PSIS-LOO ($p_{\text{loo}}$)

**Interpretation Guidelines**  
- Small → strong regularization, stable model  
- Moderate → good bias–variance trade-off  
- Large → high flexibility, possible overfitting  

Rule of thumb:  
- Compare $p_{\text{loo}}$ **relative to $n$** ($p_{\text{loo}}$/n in range [0.1 - 0.5]) and across models  

**Use in Practice**  
- Always interpret **together with $elpd_{loo}$**  
- Helps assess:
  - whether better predictive fit justifies added complexity  
- Large $p_{\text{loo}}$ + high Pareto $k$ → **LOO may be unreliable**  

**Key Takeaway**  
$p_{\text{loo}}$ = *how many parameters the model effectively uses to fit the data*  