---
title: "Normal models (with brms)"
format: html
---

# Data

As a case study, we will use (simulated) experimental data on the effects of alcohol on reaction times in a cognitive task. Forty adults were tested 20 minutes before and 20 minutes after consuming a drink. Participants were randomly assigned to one of two conditions:

- 20 participants consumed 250 ml of a 15% alcoholic beverage.
- 20 participants consumed 250 ml of a non-alcoholic beverage.

Researchers ensured that the drinks smelled and tasted the same and were served at the same temperature.

Let’s begin by loading the data.

```{r}
library(tidyverse)
theme_set(theme_classic())
alcoholRT <- read_rds("data_files/alcoholRT.rds")
alcoholRT
```

`group`: The condition to which a participant was assigned — either placebo or alcohol.
`react_time`: The difference between post-test and pre-test reaction times (in seconds). Higher values indicate worse performance in the post-test relative to the pre-test.

# Univariate Normal model

In the first part, we focus on the change in reaction time (RT) across the entire sample.

```{r}
alcoholRT %>%
  ggplot(aes(x="total sample", y=react_time)) +
  geom_violin(fill = "lightyellow") +
  geom_boxplot(fill = "lightyellow")
```

The distribution of scores appears symmetric and approximately Normal.

## Setting the prior

What do we know about the effects of alcohol on reaction time?

Most studies suggest that alcohol slows reaction times, but the magnitude of this effect depends on task complexity, duration, and other factors.

Try to set prior_mean to a reasonable value. Then set prior_sd to reflect your uncertainty about the effect of alcohol on performance. Do not confuse this value with the standard deviation of individual scores.

Plot the prior to check whether it aligns with your prior knowledge about alcohol’s effect on RT.

```{r}
# Set the values below
prior_mean
prior_sd

p_grid = seq(prior_mean - 3 * prior_sd, prior_mean + 3 * prior_sd, length.out = 100)
prior = dnorm(p_grid, mean = prior_mean, sd = prior_sd)

tibble(p_grid, prior) %>%
  ggplot(aes(p_grid, prior)) +
  geom_line() +
  labs(x = "Increase in RT (in s.) after alcohol intake", y = "Prior")
```

## Computing posterior

Before computing the posterior, we need to obtain:

- n: the number of observations
- $\bar{x}$: the sample mean
- $\sigma_0^2$: the population variance

We assume perfect knowledge of the population variance (i.e., how much individuals differ in their susceptibility to alcohol). Do not confuse this with the prior SD/variance. In practice, we would not know the true variance, but for simplicity, assume that the sample variance/SD is a perfect estimate.

```{r}
# Insert formulas to obtain required data.
n_cases = nrow(alcoholRT)
sample_mean = mean(alcoholRT$react_time)
known_sd = sd(alcoholRT$react_time)
```

To compute the posterior mean and variance for a model with known variance, use the formulas below:

$$
\hat{\mu} = \frac{\frac{m}{s^2} + \frac{n\bar{x}}{\sigma_0^2}}{\frac{1}{s^2} + \frac{n}{\sigma_0^2}}
$$



$$
\hat{\sigma}_\mu^2 = \left( \frac{1}{s^2} + \frac{n}{\sigma_0^2} \right)^{-1}
$$



Using these formulas, compute the posterior mean, variance, and standard deviation.

```{r}
# Set correct formulas.
posterior_mean
posterior_var
posterior_sd
```

Next, we visualize the results.

```{r}
posterior = dnorm(p_grid, posterior_mean, posterior_sd)
sample_df = tibble(x = sample_mean, y = dnorm(sample_mean, posterior_mean, posterior_sd))

tibble(p_grid, prior, posterior) %>%
  ggplot(aes(p_grid)) +
  geom_line(aes(y = prior, linetype = "prior")) +
  geom_line(aes(y = posterior, linetype = "posterior")) +
  geom_point(aes(x, y), data = sample_df, size = 3, colour = "red")
```

We summarize the results using the posterior median and the 95% credible interval.

```{r}
median_ci95 = c(0.025, 0.5, 0.975)
qnorm(median_ci95, posterior_mean, posterior_sd)
```

We also compute the Savage–Dickey ratio to quantify the change in evidence after observing the data.

```{r}
test_value
dnorm(test_value, posterior_mean, posterior_sd) / dnorm(test_value, prior_mean, prior_sd)
```

# Univariate Normal Model with brms

First, let’s see how the alcohol example can be implemented using `brms`.

```{r}
library(brms)
```

Below, we specify the prior in `brms` syntax:

```{r}
# You may want to change these values before moving on.
prior_mean
prior_sd
prior_text = paste0("normal(", prior_mean, ", ", prior_sd, ")")
prior_text
```

We use `set_prior` to define the prior for the mean change in reaction time (RT) after alcohol consumption. We must specify both the parameter class and the coefficient name to which the prior applies.

We then use `validate_prior` to check whether the prior has been correctly specified.

```{r}
priors = c(
  set_prior(prior_text, class = "b", coef = "Intercept")
)

validate_prior(
  prior = priors,
  formula = react_time ~ 0 + Intercept,
  data = alcoholRT
)
```

Note that in this model we also need a prior for the population standard deviation (SD), which represents how much individuals differ in their susceptibility to alcohol. We do not assume that this value is known; instead, we estimate its posterior distribution from the data. Setting informative priors for variance or SD parameters is often difficult, so `brms` provides a default weakly informative prior.

Below, we visualize this prior for $\sigma$.

```{r}
tibble(
  sigma_grid = seq(0, 8, length= 100),
  sigma_prior = dstudent_t(sigma_grid, df = 3,  mu = 0, sigma = 2.5)
) %>% 
  ggplot(aes(sigma_grid, sigma_prior))+
  geom_line()+
  labs(x = expression("Values of "~sigma), y = "Prior")
```


Next, we fit the model using `brm`. The formula `react_time ~ 0 + Intercept` indicates that reaction time is the outcome variable, and we are modeling its overall mean across the entire sample. The `0 + Intercept` notation tells brms that there are no predictors — we are only estimating the intercept.

```{r}
fit_normal <- brm(
  formula = react_time ~ 0 + Intercept,
  data = alcoholRT,
  prior = priors,
  sample_prior = "yes"
)
```

We can inspect the results with:

```{r}
fit_normal
```

To compute the Savage–Dickey ratio, we use the `hypothesis` function:

```{r}
hypothesis(fit_normal, "Intercept = 0")
```

We can also visualize the result:

```{r}
hypothesis(fit_normal, "Intercept = 0") %>%
  plot()
```

Compare these results with the analytic solution.


# Two-Group Normal Model with brms

Start by specifying priors for the change in reaction time (RT) in the alcohol and placebo groups.

```{r}
# You may want to change these values before moving on.
prior_mean_alcohol
prior_sd_alcohol

prior_mean_placebo
prior_sd_placebo

prior_text_alcohol = paste0("normal(", prior_mean_alcohol, ", ", prior_sd_alcohol, ")")
prior_text_placebo = paste0("normal(", prior_mean_placebo, ", ", prior_sd_placebo, ")")
prior_text_alcohol
prior_text_placebo
```

Set priors for both groups:

```{r}
priors = c(
  set_prior(prior_text_alcohol, class = "b", coef = "groupalcohol"),
  set_prior(prior_text_placebo, class = "b", coef = "groupplacebo")
)

validate_prior(
  prior = priors,
  formula = react_time ~ 0 + group,
  data = alcoholRT
)
```

Fit the model. 
The formula `react_time ~ 0 + group` specifies that the outcome (`react_time`) is predicted by a categorical variable (`group`). The `0 + group` notation ensures that we estimate separate means for both conditions. Without `0 +`, `brms` would estimate one group mean and the difference between groups.

```{r}
fit_normal_two_group <- brm(  
  formula = react_time ~ 0 + group,
  data = alcoholRT,
  prior = priors,
  sample_prior = "yes"
)
```

Inspect the results:

```{r}
fit_normal_two_group
```

To compare group means, we can use the `hypothesis` function:

```{r}
hypothesis(fit_normal_two_group, "groupalcohol - groupplacebo = 0")
```

We can also visualize the posterior distribution of the difference:

```{r}
hypothesis(fit_normal_two_group, "groupalcohol - groupplacebo = 0") %>% 
  plot()
```

Finally, compare these results with a frequentist approach:

```{r}
t.test(react_time ~ group, data = alcoholRT, var.equal=T)
```




