---
title: "Binary Classification"
format: html
---

## Vampire problem

In the population of 100,000 individuals, 100 are vampires.

$P(vampire = positive)$
Proportion of population affected by the condition.

```{r}
(prevalence) 
```

There is a test for vampirism construed by some big-pharma company, but...
- Of the 100 who are vampires, only 80 of them will test positive for vampirism.

$P(test = positive | vampire = positive)$
(True positive rate) Proportion of those who are affected by the condition and are correctly tested positive.

```{r}
(sensitivity) 
```

- Of the 99,900 mortals, 4995 of them will test positive for vampirism.

$P(test = negative | vampire = negative)$
(True negative rate) Proportion of those who are not affected by the condition and are correctly tested negative.

```{r}
(specificity)
```

$P(test = positive | vampire = negative)$
Complement proportion to specificity.

```{r}
(false_positive_rate)
```

$P(test = negative | vampire = positive)$
Complement proportion to sensitivity.

```{r}
(false_negative_rate)
```

What is a probability that someone who obtained a positive test is a vampire?

$P(vampire = positive ∧ test = positive)$
Proportion of a population affected by a condition and correctly tested positive.

```{r}
(true_positive)
```

$P(vampire = negative ∧ test = positive)$
Proportion of a population not affected by a condition and incorrectly tested positive.

```{r}
(false_positive)
```

$P(vampire = negative ∧ test = negative)$
Proportion of a population not affected by a condition and correctly tested negative.

```{r}
(true_negative)
```

$P(vampire = positive ∧ test = negative)$
Proportion of a population affected by a condition and incorrectly tested negative.

```{r}
(false_negative)
```


$P(vampire = positive | test = positive)$
Proportion of those who tested positive and are affected by the condition.

```{r}
(positive_predictive_value)
```

$P(vampire = negative | test = negative)$
Proportion of those who tested negative and are not affected by the condition.

```{r}
(negative_predictive_value)
```

$P(vampire = negative | test = positive)$
Proportion of false positives in the pool of those that test positive.

```{r}
(false_discovery_rate)
```

$P(vampire = positive | test = negative)$
Proportion of false negatives in the pool of those that test negative.

```{r}
(false_omission_rate)
```


$P(vampire = positive ∧ test = positive ∨ vampire = negative ∧ test = negative)$
Proportion of the population that is true positive or true negative.

```{r}
(accuracy)
```


